A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-06-24 sections/01DataDescription.Rmd
typos

The data set we use consists of three studies that followed migrants who had recently arrived in the Netherlands in their daily interactions with the Dutch majority group members \citet[][]{Kreienkamp2022b}. After a general migration-focused pre-questionnaire, participants were invited twice per day to report on their (potential) interactions with majority group members for at least 30 days. The short ESM surveys were sent out at around lunch (12pm) and dinner time (7pm). After the 30 day study period participants filled in a post-questionnaire that mirrored the pre-questionnaire. Participants received either monetary compensation or partial course credits based on the number of surveys they completed.

The original studies included `r sum(missS123$pptFull)` participants ($N_{S1}=$ `r missS123$pptFull[missS123$study==1]`, $N_{S2}=$ `r missS123$pptFull[missS123$study==2]`, $N_{S3}=$ `r missS123$pptFull[missS123$study==3]`) with a total of `r format(sum(missS123$nFull), big.mark=",")` ESM measurements. Each of the studies focused on recently arrived first-generation migrants and each study included a number of ideosyncratic variables relevant for the broader research collective. For our empirical example we focus on the variables that were collected during the ESM surveys and were available in all three studies. Variable selection and preparation is described as part of the illustration below but additional methodological details about the study setup are available in \citet[][]{Kreienkamp2022b}.

>>>>

The data set we use consists of three studies that followed migrants who had recently arrived in the Netherlands in their daily interactions with the Dutch majority group members \citep[for the data set see][]{Kreienkamp2022b}. After a general migration-focused pre-questionnaire, participants were invited twice per day to report on their (potential) interactions with majority group members for at least 30 days. The short ESM surveys were sent out at around lunch (12pm) and dinner time (7pm). After the 30 day study period participants filled in a post-questionnaire that mirrored the pre-questionnaire. Participants received either monetary compensation or partial course credits based on the number of surveys they completed.

The original studies included `r sum(missS123$pptFull)` participants ($N_{S1}=$ `r missS123$pptFull[missS123$study==1]`, $N_{S2}=$ `r missS123$pptFull[missS123$study==2]`, $N_{S3}=$ `r missS123$pptFull[missS123$study==3]`) with a total of `r format(sum(missS123$nFull), big.mark=",")` ESM measurements. Each of the studies focused on recently arrived first-generation migrants and each study included a number of ideosyncratic variables relevant for the broader research collective. For our empirical example we focus on the variables that were collected during the ESM surveys and were available in all three studies. Variable selection and preparation is described as part of the illustration below but for additional methodological details about the study setup see \citet[][]{Kreienkamp2022b}.

### 2023-06-24 sections/03FeatureExtraction.Rmd
typos

\paragraph{Variability.}
Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. In time series analyses, variability is conceptually important because information about the distribution and diversity of the data has been found to be indicative of worse psychological states \citep{myin-germeys2018}. Person level differences of ESM measurements have, for example, been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. As such, psychological researchers and practitioners are often empirically interested in person-level differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or vary across the measurement period?''

>>>>

\paragraph{Variability.}
Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. In time series analyses, variability is conceptually important because information about the distribution and diversity of the data has been found to be indicative of worse psychological states \citep{myin-germeys2018}. Person level differences of ESM measurements have, for example, been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. As such, psychological researchers and practitioners are often empirically interested in person-level differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or do they vary across the measurement period?''

### 2023-06-24 sections/07ClusterEvaluationInterpretation.Rmd
:memo: feat red adj
typos

We first inspect the clusters based on the average values of meaningful features \citep[see \fgrref[A]{fig:clusterFeatVar}][]{Kennedy2021}. We (1) see that for some variables the features are generally stronger in separating the clusters (e.g., 'how cooperative an interaction was' compared to 'attitudes towards the Dutch'). Additionally, we see that (2) within the variables some features are better at distinguishing clusters (e.g., median of well-being vs. MAC of well-being).We then inspect the clusters with a focus on the features (see \fgrref[B]{fig:clusterFeatVar}). While this is the same data as for the variable focus, we can see more clearly that some features are better at distinguishing the clusters across variables (e.g., mean and median compared to auto correlations). This offers some information on which features were are most important in understanding the two extracted groups.

@@

In the second step, we look at prototypical trajectories of the clusters. For k-means clustering it is often recommended to use the average over time of the responses within the cluster \citep[see \fgrref{fig:clusterTs}][]{niennattrakul2007}\footnote{It is important to note, however, that direct comparability can be a concern, and often times some subset selection or nonlinear alignment is necessary \citep[e.g.,][]{gupta1996}. Additionally, finding cluster prototypes is often substantially easier with embedded clustering methods because in many cases a cluster-level model is estimated as part of the expectation–maximization procedure \citep[e.g.,][]{denteuling2021}. For medoid-based clustering algorithms, a common approach is simply using cluster medoid as the prototype \citep{kaufman1990}.}. Immediately striking are the mean differences, where participants in the second cluster had more meaningful and fulfilling outgroup interactions also consistently reported more voluntary and cooperative interactions but less accidental and involuntary interactions. The same cluster also reported an increase in need fulfilling interaction over the 30 day period and an increase in interactions that were representative of the outgroup. Whereas the other cluster showed a decrease in voluntary, cooperative, and positive interactions over the 30 days. This 'deterioration' cluster also saw a decrease in general need fulfillment and experienced well-being over the 30 days (see \fgrref[B]{fig:clusterTs}). We also see that while interaction representativeness, outgroup attitudes, well-being are relatively stable for both clusters, the deteriorating cluster also showed substantially higher variablity and instability over time (also see \fgrref[A]{fig:clusterTs}).

@@

Finally, we can also assess the clusters across other person-level variable \citep[e.g.,][]{monden2022}. This out-of-feature comparison allows us to check for data artifacts, as well as check whether the developmental clusters are associated with important social markers and individual differences. To illustrate artifact checks, we added the number of measurements into the comparison and find that the participants in the detereoration cluster on average completed more ESM surveys and reported on more intergroup interactions than the cluster with the more positive interactions (see \fgrref[B]{fig:clusterTs}). While this difference could indicate that the clusters might not entirely be comparable in the response patterns, we can find some relief in our data exclusion procedures during which we ensured that the general time frame and completion rates were not too dissimilar. To illustrate the utility of individual differences, we compare the two samples in terms of the participants' self-reported discrimination experiences in the Netherlands (measured during the post-measurement). \fgrref[B]{fig:clusterTs} illustrates that participants in the deteriorating cluster reported substantially higher levels of everyday discrimination. Thus, both intensive longitudinal and cross-sectional variables that were not included in the original clustering step can be used to explore and understand the cluster differences in more detail.

>>>>

We first inspect the clusters based on the average values of meaningful features (see \fgrref[A]{fig:clusterFeatVar}; \citealp{Kennedy2021}). We (1) see that for some variables the features are generally stronger in separating the clusters (e.g., 'how cooperative an interaction was' compared to 'attitudes towards the Dutch'). Additionally, we see that (2) within the variables some features are better at distinguishing clusters (e.g., median of well-being vs. MAC of well-being). We then inspect the clusters with a focus on the features (see \fgrref[B]{fig:clusterFeatVar}). While this is the same data as for the variable focus, we can see more clearly that some features are better at distinguishing the clusters across variables (e.g., mean and median compared to auto correlations). This offers some information on which features were are most important in understanding the two extracted groups.

@@

In the second step, we look at prototypical trajectories of the clusters. For k-means clustering it is often recommended to use the average over time of the responses within the cluster \citep[see \fgrref{fig:clusterTs}][]{niennattrakul2007}\footnote{It is important to note, however, that direct comparability can be a concern, and often times some subset selection or nonlinear alignment is necessary \citep[e.g.,][]{gupta1996}. Additionally, finding cluster prototypes is often substantially easier with embedded clustering methods because in many cases a cluster-level model is estimated as part of the expectation–maximization procedure \citep[e.g.,][]{denteuling2021} or S-GIMME \citep[e.g.][]{lane2019}. For medoid-based clustering algorithms, a common approach is simply using cluster medoid as the prototype \citep{kaufman1990}.}. Immediately striking are the mean differences, where participants in the second cluster had more meaningful and fulfilling outgroup interactions also consistently reported more voluntary and cooperative interactions but less accidental and involuntary interactions. The same cluster also reported an increase in need fulfilling interaction over the 30 day period and an increase in interactions that were representative of the outgroup. Whereas the other cluster showed a decrease in voluntary, cooperative, and positive interactions over the 30 days. This 'deterioration' cluster also saw a decrease in general need fulfillment and experienced well-being over the 30 days (see \fgrref[B]{fig:clusterTs}). We also see that while interaction representativeness, outgroup attitudes, well-being are relatively stable for both clusters, the deteriorating cluster also showed substantially higher variablity and instability over time (also see \fgrref[A]{fig:clusterTs}).

@@

Finally, we can also assess the clusters across other person-level variable \citep[e.g.,][]{monden2022}. This out-of-feature comparison allows us to check for data artifacts, as well as check whether the developmental clusters are associated with important social markers and individual differences. To illustrate artifact checks, we added the number of measurements into the comparison and find that the participants in the detereoration cluster on average completed more ESM surveys and reported on more intergroup interactions than the cluster with the more positive interactions (see \fgrref[B]{fig:clusterTs}). While this difference could indicate that the clusters might not entirely be comparable in the response patterns, we can find some relief in our data exclusion procedures during which we ensured that the general time frame and completion rates were not too dissimilar. To illustrate the utility of individual differences, we compare the two samples in terms of the participants' self-reported discrimination experiences in the Netherlands (measured during the post-measurement). \fgrref[B]{fig:clusterFeatVar} illustrates that participants in the deteriorating cluster reported substantially higher levels of everyday discrimination. Thus, both intensive longitudinal and cross-sectional variables that were not included in the original clustering step can be used to explore and understand the cluster differences in more detail.
